---
title: "Methylation pattern"
output:
 html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
date: "2024-09-11"
---

# Setup
```{r setup, include=FALSE}
# Set printing source code
knitr::opts_chunk$set(echo = TRUE)

# Load packages
# CRAN packages
library(tidyverse)
library(patchwork)
library(gridExtra)
library(lme4)
library(lmerTest)
library(here)
library(NbClust)
library(ComplexHeatmap)
library(emmeans)
library(performance)
library(see)
library(glmmTMB)
library(tibble)
library(openxlsx)


#Bioconductor packages
library(GenomicRanges)
library(GenomicFeatures)
library(PCAtools)
library(Biostrings)

select = dplyr::select

# Set correct select function
select = dplyr::select

# Define tissue colors
types = c("Stipe","Lamella","Spore","control")
colors = c("#96C176FF","#CF9DDBFF","#4472C4FF","#000000FF")
types = c("Stipe","Lamella","Spore","control","stipe","lamella","spore") # Catch lowercase tissues
colors = c("#96C176FF","#CF9DDBFF","#4472C4FF","#000000FF","#96C176FF","#CF9DDBFF","#4472C4FF")
names(colors) = types

# Define genomic context colors
annotationColors = c("mRNA" = "#4472C4","TE" = "#C00000","Both" = "#9D475B","intergenic" = "#AFABAB")

# Define methylation level colors (< 0.05, 0.05 - 0.30, 0.30 - 0.75, 0.75 - 0.90, > 0.90)
methylationLevelColors = my_colors <- c(
  "Nothing"   = "#4575B4",
  "Low"  = "#91BFDB",
  "Medium"  = "#E0F3F8",
  "High"  = "#FEE090",
  "All" = "#D73027"
)

```




# Annotation

### TE reference
```{r, message=FALSE}
# Define GFF3 columns
gff3Colnames =c("seqid", "source",
                "type","start",
                "end", "score",
                "strand","phase",
                "attributes")

  # Load information about TE classes, included in repository
  TE.classes = read_csv(file = here("annotation","Repeat.classes.csv")) %>%
    rename(Name = "TEfamily") %>% 
    rename("Superfamily / Notes" = "Notes") #columns renamed for downstream use

  # Define TE file location, included in repository
  TEFile <- here("annotation","Maror2.repeatmasker.rmlib2.3.gff")

  # Read in TE dataframe from gff file
  TEdf =    read_tsv(file = TEFile, col_names = gff3Colnames,comment = "#") %>% separate(attributes, into = c("ID","target"), sep = ";") %>%
    mutate(ID = as.numeric(gsub("ID=","",ID))) %>%
    separate(target, into = c("Target","motif",
                              "motifStart","motifStop"),
             sep = " ") %>%
    mutate(motif = gsub("\"","",motif))%>%
    mutate(motif = gsub("Motif:","",motif))  %>%
    select(seqid,start,end,score,
           strand, ID,motif,
           motifStart,motifStop) %>%
    rename(ID = "TEID") %>%
    mutate(TE_ID2 = 1:n()) %>%
    mutate(motifStart = as.numeric(gsub(" ","",motifStart))) %>%
    mutate(motifStop = as.numeric(gsub(" ","",motifStop))) %>%
    dplyr::rename(TEclass = "motif") %>%
    mutate(TEclass = gsub(" ","",TEclass)) %>%
    mutate(TEclass = gsub("Maror","",TEclass)) 

 
  # Generate GRanges of TE file
  TEs = makeGRangesFromDataFrame(TEdf,
                                 seqnames.field = "seqid", 
                                 start.field = "start", 
                                 end.field = "end",
                                 strand.field = "strand",
                                 keep.extra.columns = TRUE)
  

# Merge TE classes with dataframe to add annotation and remove other repeat annotations
# From 13309 to 8893 repeats
TEdf%>%
  separate(TEclass, into = c("TEfamily","nr"), sep = "-",remove = FALSE) %>%
  inner_join(TE.classes)->
  TEdf

```


## Gene reference 
```{r, message=FALSE}

gtfFile = here("annotation","GCF_018924745.1_UU_Maror_2_genomic.gff")	# Included in repository
txdb <- makeTxDbFromGFF(file=gtfFile, dataSource="ensemblgenomes",format = "gff3")

# define GFF3 columns
gff3Colnames =c("seqid", "source",
                "type","start",
                "end", "score",
                "strand","phase",
                "attributes")
  # Read in the gff file
  annotation.df =    read_tsv(file = gtfFile, col_names = gff3Colnames,comment = "#") 

# Keep only mRNA annotations
 annotation.df %>% filter(type =="mRNA") %>%
   group_by(seqid,start,end, strand) %>%
   slice_head(n = 1) %>% # Remove multiple mRNA annotations on the same genomic region
   ungroup() %>%
   separate_longer_delim(attributes, delim = ";") ->
   test
  # Keep only relevant attributes and give them separate columns
  # Drop any unused columns
  test %>% select(seqid,start,end,strand,attributes) %>%
   separate(attributes, into = c("key","value"), sep = "=") %>%
   filter(key %in% c("Name","locus_tag","product")) %>%
   pivot_wider(values_from = "value",names_from = "key") -> 
   annotation.mRNA.df

# Add tx_id column
annotation.mRNA.df %>% select(seqid,start,end, strand,Name,locus_tag,product) %>%
  mutate(tx_id = 1:n()) -> 
  annotation.mRNA.df

# Create GRanges object
mRNAs = makeGRangesFromDataFrame(annotation.mRNA.df,
                                 seqnames.field = "seqid", 
                                 start.field = "start", 
                                 end.field = "end",
                                 strand.field = "strand",
                                 keep.extra.columns = TRUE)
  

```

Run only once for getting relationship between protein ID (XP) and gene ID (XM)
```{r, inlcude false  }
# Add XM and XP IDs
gene2ProteinFile = here("annotation","gene2protein_relationship.csv")

relationship_df = read_csv(gene2ProteinFile)

relationship_df %>% mutate(Name = XM_ID) -> relationship_df
annotation.mRNA.df = inner_join(annotation.mRNA.df,relationship_df , by = "Name")
```


# Samples 
```{r, message=FALSE}
# Read and organize sample data
samples.ring.methyl.df = read_csv(file = here("data","meDNA","meDNA.metaData.csv"))

samples.ring.methyl.df %>% arrange(Tissue,Location,Replicate) -> 
  samples.ring.methyl.df
```

# Read in data

Loading it from the google drive. This might change when we publish

```{r, message=FALSE}

#TODO: Insert DOI
# Data can be obtained from ###INSERT DOI#####
meData_path <- here("data", "meDNA", "methylationData.samples.QC.csv.gz")

# Read in the data, read_csv automatically handles the decompression
# Data consists of every position of every sample that is considered in the final dataset. 
methylationData.3.QC <- read_csv(meData_path)

```

#Figure 3
###  TE analysis

```{r }
# Filter methylation information on TE locations
TEdf %>% 
  mutate(motifStart = as.numeric(motifStart)) %>%
  mutate(motifStop = as.numeric(motifStop)) %>%
  separate(TEclass, into=c("TEclass","nr"), sep = "-") ->
  TEdf

# Extract TE info
TEdf %>% select(TEclass,seqid,TE_ID2) -> 
  TE.info

# inner join on TE_ID2
methylationData.3.QC %>% 
  select(-TEclass) %>%
  filter(location == "TE") %>%
  inner_join(TE.info) ->
  methylationData.3.QC.TE ##5649 TEs
```

### Data filtering

```{r }
# Extract methylation percentage of all TEs per sample
methylationData.3.QC.TE %>%
  group_by(TE_ID2,Location,Replicate2,Tissue) %>%
  summarise( percent = mean(percent)) ->
  methylationData.3.QC.TE.summary 

# extract by how many samples a TE is covered per Tissue
methylationData.3.QC.TE.summary %>%
  group_by(TE_ID2,Tissue) %>%
  summarise(nrOfExamples = n()) -> methylationData.3.QC.TE.summary.2

# Remove TEs that are not included in all 3 of a mushroom
methylationData.3.QC.TE.summary.2 %>%
  group_by(Tissue) %>%
  summarise(n = max(nrOfExamples)) %>%
  inner_join(methylationData.3.QC.TE.summary.2) %>%
  group_by(TE_ID2) %>%
  summarise(n = n()) %>%
  filter(n == 3)  %>%
  select(TE_ID2)->
  TE.covered ##5616 TEs

# And remove TEs that are not covered by all samples
methylationData.3.QC.TE.summary %>%
  inner_join(TE.covered) %>%
  pivot_wider(values_from = "percent",names_from = c("Location","Replicate2","Tissue")) %>%
  column_to_rownames(var = "TE_ID2")%>%
  drop_na()->                     # Removes any row with at least one NA->
  methylationData.3.QC.TE.wide ##3843 TEs

```

#### Filtering on TEs with sufficient informative sites (>10 CpG sites) and covered by at least 2 tissues
```{r}
# Now we test for every TE if it is differentially methylated by location, tissue or replicate.
# Extract unique TEs in methylation data: 5649 of 8893 TEs included
methylationData.3.QC.TE %>% distinct(TE_ID2) ->
  TE.unique

# detemerine for every tissue the methylation level at each TE start position
methylationData.3.QC.TE %>%
  select(TE_ID2,Sample,Tissue,start,percent) %>%
  ungroup() %>%
  group_by(TE_ID2,Tissue,start) %>%
    summarise(n = n(), methylation = mean(percent)) ->
  TE.summary.all

# Filter out TEs with few CpG sites (<10) and those only covered in 1 tissue 
TE.summary.all %>%
  group_by(TE_ID2,Tissue) %>%
  summarise(n = n()) %>%
  ungroup %>%
  filter(n >=10) %>%
  group_by(TE_ID2) %>%
  summarise(n = n()) %>%
  filter(n >=2) ->
  TE.filtered # 3185 TEs

# Filter to select only TEs with >10 unique positions covered and count how many positions they cover
TE.summary.all %>% filter(TE_ID2 %in% TE.filtered$TE_ID2) %>%
  ungroup() %>%
  distinct(TE_ID2,start) %>%
  group_by(TE_ID2) %>%
  summarise(n = n()) %>%
  filter(n > 10) -> # Consider changing to less stringent filter
  TE.filtered.nCpG.nObs # Final TEs tested 2965

```

### Running linear mixed model

```{r, message = FALSE, warning = FALSE}

# lmer for Tissue
dist.list.Tissue = list() # create empty list
for(TE in TE.filtered.nCpG.nObs$TE_ID2){
  TEinfo = list() # create list to store result
    methylationData.3.QC.TE %>%
      filter(TE_ID2 == TE) %>% # Filter on TE of interest
      select(Sample,Tissue,start,percent) ->
      mat #store relevant data
     model <- lmer(percent ~ Tissue + (1 | start), data = mat) # perform linear ME  model, with start as random intercept
       pval  <- anova(model)$`Pr(>F)`[1]
     TEinfo[["Name"]] = TE
     TEinfo[["data"]] = mat
     TEinfo[["mixedLinearModel"]] = model
     TEinfo[["pvalue"]] = pval
     
     dist.list.Tissue[[TE]] = TEinfo # Store result
       
}

# lmer for Location
dist.list.Location = list()
for(TE in TE.filtered.nCpG.nObs$TE_ID2){
  TEinfo = list()
    methylationData.3.QC.TE %>%
      filter(TE_ID2 == TE) %>%
      select(Sample,Location,start,percent) ->
      mat
     model <- lmer(percent ~ Location + (1 | start), data = mat)
       pval  <- anova(model)$`Pr(>F)`[1]
     TEinfo[["Name"]] = TE
     TEinfo[["data"]] = mat
     TEinfo[["mixedLinearModel"]] = model
     TEinfo[["pvalue"]] = pval
     
     dist.list.Location[[TE]] = TEinfo
       
}

# lmer for Replicate
dist.list.Rep = list()
for(TE in TE.filtered.nCpG.nObs$TE_ID2){
  TEinfo = list()
    methylationData.3.QC.TE %>%
      filter(TE_ID2 == TE) %>%
      select(Sample,Replicate2,start,percent) ->
      mat
     model <- lmer(percent ~ Replicate2 + (1 | start), data = mat)
       pval  <- anova(model)$`Pr(>F)`[1]
     TEinfo[["Name"]] = TE
     TEinfo[["data"]] = mat
     TEinfo[["mixedLinearModel"]] = model
     TEinfo[["pvalue"]] = pval
     
     dist.list.Rep[[TE]] = TEinfo
       
}

# Extract p-values
TEspvalues.Tissue = list()
for(TE in TE.filtered.nCpG.nObs$TE_ID2){
  data.frame(
  pvalue = dist.list.Tissue[[TE]][["pvalue"]],
  TE_ID2 = dist.list.Tissue[[TE]][["Name"]],
  Test = "Tissue"
  ) ->
  TEspvalues.Tissue[[as.character(TE)]]
}

TEspvalues.Location = list()
for(TE in TE.filtered.nCpG.nObs$TE_ID2){
  data.frame(
  pvalue = dist.list.Location[[TE]][["pvalue"]],
  TE_ID2 = dist.list.Location[[TE]][["Name"]],
  Test = "Location"
  ) ->
  TEspvalues.Location[[as.character(TE)]]
}

TEspvalues.Rep = list()
for(TE in TE.filtered.nCpG.nObs$TE_ID2){
  data.frame(
  pvalue = dist.list.Rep[[TE]][["pvalue"]],
  TE_ID2 = dist.list.Rep[[TE]][["Name"]],
  Test = "Rep"
  ) ->
  TEspvalues.Rep[[as.character(TE)]]
}


# Adjust p values and extract significant at p.adjust < 0.0001
bind_rows(TEspvalues.Tissue) %>%
  mutate(adj_pvalue = p.adjust(pvalue, method = "BH")) %>%
  mutate(significant = "No") %>%
  mutate(significant = replace(significant, adj_pvalue < (0.0001), "Yes")) ->
  lmm_results.TE.Tissue


bind_rows(TEspvalues.Location) %>%
  mutate(adj_pvalue = p.adjust(pvalue, method = "BH")) %>%
  mutate(significant = "No") %>%
  mutate(significant = replace(significant, adj_pvalue < (0.0001), "Yes")) ->
  lmm_results.TE.Location

bind_rows(TEspvalues.Rep) %>%
  mutate(adj_pvalue = p.adjust(pvalue, method = "BH")) %>%
  mutate(significant = "No") %>%
  mutate(significant = replace(significant, adj_pvalue < (0.0001), "Yes")) ->
  lmm_results.TE.Rep

# Combine results
lmm_results.TE = bind_rows(lmm_results.TE.Tissue,lmm_results.TE.Location,lmm_results.TE.Rep )

# Plot results of linear mixed model

lmm_results.TE %>% ggplot(aes(x = pvalue, color = Test)) +
  geom_density( adjust = 1/10) +
  labs(
    title = "Density plot",
    x = "p-value",
    y = "Density",
    shape = "Class"
  ) + theme_light() ->
  TE.pvalue.density.plot

ggsave(here("figures","Figure3","Fig3A.pdf"),
       plot = TE.pvalue.density.plot,width = 180, height = 180, units = "mm", dpi = 300)

# Extract significance values

lmm_results.TE %>% arrange(adj_pvalue) %>%
  filter(significant == "Yes") ->
  lmm_results.TE.significant
lmm_results.TE.significant %>% 
  group_by(Test) %>%
  summarise(numberOfSignificant = n()) ->
  lmm_results.TE.significant.summary

```

###  Aalysis of  significant TEs  
```{r }  
# extract only those TEs initially included in the analysis  
lmm_results.TE.Tissue %>% filter(significant == "Yes") %>%
  arrange(adj_pvalue) ->
  lmm_results.TE.Tissue.significant

TE_tissue_DE = intersect(lmm_results.TE.Tissue.significant$TE_ID2,as.numeric( rownames(methylationData.3.QC.TE.wide))) 

methylationData.3.QC.TE.wide[as.character(TE_tissue_DE),]->
methylationData.3.QC.TE.sig.wide 

```
#### Cluster of TEs

##### Hierarchical clustering with correct number of clusters
```{r}
## Cluster analysis. Determine optimal number of clusters and assign each TE

dist_matrix <- dist(methylationData.3.QC.TE.sig.wide, method = "euclidean")
hc <- hclust(dist_matrix, method = "ward.D2")
nb <- NbClust(methylationData.3.QC.TE.sig.wide, distance="euclidean", min.nc=2, max.nc=10, method="ward.D2")


wss <- numeric(10)
for (k in 2:10) {
  clusters <- cutree(hc, k)
  wss[k] <- sum(sapply(unique(clusters), function(c) {
    pts <- methylationData.3.QC.TE.sig.wide[clusters == c, ]
    sum(scale(pts, scale=FALSE)^2)
  }))
}

plot(1:10, wss, type="b", xlab="k", ylab="Within-cluster SS")




nrOfClusters = nb$Best.nc[[1]]

nrOfClusters = 3
clusters <- cutree(hc, nrOfClusters)


clusters %>%  as.data.frame() %>%
  rownames_to_column(var = "TE_ID2") %>%
  rename("." = "cluster") ->
  clusters




```


#####  Visualisation of 3 clusters

```{r}

#Figure 3C
# Note orders have been manually set to fit the heatmap

clusters %>%
  mutate(TE_ID2 = as.double(TE_ID2)) %>%
  inner_join(methylationData.3.QC.TE) %>%
  group_by(cluster, Tissue) %>%
  summarise(mean = mean(percent), sd = sd(percent))->
  clusters.Tissue.summary


clusters.Tissue.summary %>% group_by(cluster) %>%
  summarise(overallMean = mean(mean)) %>%
  arrange(overallMean) %>%
  mutate(clusterOrder = as.factor(row_number()))->
  cluster.order
  

clusters %>%  
  mutate(TE_ID2 = as.double(TE_ID2)) %>%
  inner_join(methylationData.3.QC.TE) %>%
  ungroup() %>%
  distinct(TE_ID2,cluster,TEclass) %>%
  group_by(cluster,TEclass) %>%
  summarise(nrOfTEs = n()) %>%
  arrange(cluster, -nrOfTEs)%>%
  group_by(cluster) %>%
  slice_head(n = 5)->
  clusters.count.summary


clusters.Tissue.summary %>%
  mutate(clusterName = "Low Methylation") %>%
  mutate(clusterName = replace(clusterName, 
                               grepl(1, cluster),
                               "Medium Methylation")) %>% 
  
  mutate(clusterName = replace(clusterName, 
                               grepl(3, cluster),
                               "High Methylation")) %>%
  mutate(clusterName = factor(clusterName, 
                              levels = c("Low Methylation",
                                         "Medium Methylation",
                                         "High Methylation"))) %>%
  inner_join(clusters.count.summary)->
clusters.summary

clusters.Tissue.summary %>%
  inner_join(clusters.count.summary) %>%
  inner_join(cluster.order)->
clusters.summary
  
ha = HeatmapAnnotation(
    cluster = as.factor(clusters$cluster)
)


pdf(here("figures", "Figure3", "Fig3B.pdf"),
    width = 8, height = 6)
ComplexHeatmap::Heatmap(t(methylationData.3.QC.TE.wide[as.character(TE_tissue_DE),]),
                        show_column_names = FALSE,
                        show_column_dend = TRUE, 
                        cluster_columns = hc,
                        column_split = nrOfClusters,
                        top_annotation = ha,
                        cluster_rows =TRUE,  
                        split = 2
) 

dev.off()

clusters %>%
  mutate(TE_ID2 = as.double(TE_ID2)) %>%
  inner_join(methylationData.3.QC.TE) %>%
  group_by(TE_ID2,Tissue,Sample,cluster ) %>%
  summarise(percent = mean(percent) )%>%
  inner_join(clusters.summary) %>%
  ggplot(aes(x = clusterOrder, y = percent, fill = Tissue)) +
  geom_boxplot() +
  theme_light() +
  labs(
    title = "Methylation levels per TE cluster",
    x = "TE Cluster",
    y = "Percent Methylation"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),  # Rotate and adjust x-axis labels
    axis.title.x = element_text(size = 12, face = "bold"),  # Format x-axis title
    axis.title.y = element_text(size = 12, face = "bold")   # Format y-axis title
  )+
  scale_fill_manual(values = colors) -> distributionPerClusterPlot
   
ggsave(here("figures","Figure3","Fig3C.pdf"),
       plot = distributionPerClusterPlot ,width = 8, height = 6, units = "in", dpi = 300)    

```

## Significance testing
```{r}
# Get counts per class of TEs per cluster
clusters %>%
  mutate(TE_ID2 = as.double(TE_ID2)) %>%
  inner_join(TEdf) %>%
  count(cluster, Class) %>%
  pivot_wider(names_from = Class, values_from = n, values_fill = 0) -> cluster_summary

# Perform chi-squared test to test significance
#Then perform a chi-squared
count_matrix <- cluster_summary %>% 
  select(`I (Retro)`, `II (DNA)`) %>% 
  as.matrix()
chisq_results <- chisq.test(count_matrix)
print(chisq_results)
chisq_results$residuals

# Test significance of Figure 3C
# Extract percentage methylation per tissue

# Note: Cluster 1 middle, cluster 2 is low and cluster 3 is high here
# In all three comparisons lamellae vs spores p > 0.05, while stipe vs anything p < 0.0001
clusters %>%
  mutate(TE_ID2 = as.double(TE_ID2)) %>%
  inner_join(methylationData.3.QC.TE) %>%
  group_by(TE_ID2,Tissue,cluster ) %>%
  summarise(percent = mean(percent) )-> clusters.methylation.percent

# Middle methylation levels
cluster_1_data <- clusters.methylation.percent %>% filter(cluster == 1)
cluster_1_data <- cluster_1_data %>%
  mutate(prop = percent / 100)
n <- nrow(cluster_1_data)
cluster_1_data$prop_adj <- (cluster_1_data$prop * (n - 1) + 0.5) / n # Adjust data to enable a beta model

beta_mod_1 <- glmmTMB(prop_adj ~ Tissue + (1|TE_ID2), 
                    data = cluster_1_data, 
                    family = beta_family())

tissue_comp_1 <- emmeans(beta_mod_1, pairwise ~ Tissue, type = "response")
tissue_comp_1$contrasts

# Low methylation levels
cluster_2_data <- clusters.methylation.percent %>% filter(cluster == 2)
cluster_2_data <- cluster_2_data %>%
  mutate(prop = percent / 100)
n <- nrow(cluster_2_data)
cluster_2_data$prop_adj <- (cluster_2_data$prop * (n - 1) + 0.5) / n # Adjust data to enable a beta model

beta_mod_2 <- glmmTMB(prop_adj ~ Tissue + (1|TE_ID2), 
                    data = cluster_2_data, 
                    family = beta_family())

tissue_comp_2 <- emmeans(beta_mod_2, pairwise ~ Tissue, type = "response")
tissue_comp_2$contrasts

# high methylation levels
cluster_3_data <- clusters.methylation.percent %>% filter(cluster == 3)
cluster_3_data <- cluster_3_data %>%
  mutate(prop = percent / 100)
n <- nrow(cluster_3_data)
cluster_3_data$prop_adj <- (cluster_3_data$prop * (n - 1) + 0.5) / n # Adjust data to enable a beta model

beta_mod_3 <- glmmTMB(prop_adj ~ Tissue + (1|TE_ID2), 
                    data = cluster_3_data, 
                    family = beta_family())

tissue_comp_3 <- emmeans(beta_mod_3, pairwise ~ Tissue, type = "response")
tissue_comp_3$contrasts



```

## Supplementary table 2

```{r}
#here I create a supplementary table based on the significance test of all the TEs included in the analysis
#conver TE_ID2 to an integer
publication_table <- lmm_results.TE

#join TE annotation
publication_table <- merge(publication_table, TEdf, by="TE_ID2")

#prepare methylation_data
methylationData.3.QC.TE.wide.TEID <- rownames_to_column(methylationData.3.QC.TE.wide)
names(methylationData.3.QC.TE.wide.TEID)[names(methylationData.3.QC.TE.wide.TEID) == "rowname"] <- "TE_ID2"

#Attempt the merge
publication_table <- merge(publication_table, methylationData.3.QC.TE.wide.TEID, by="TE_ID2")

#Export to excel
write.xlsx(publication_table, here("results", "meDNA", "Supplementary_table_meDNA.xlsx"))

```